---
title: "HR Analytics"
author: "Ishwarya keerthivasan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Comprehensive HR Analytics Portfolio: HR Employee Attrition Analysis**

## **1. Data Understanding and Preprocessing**

### **Load Libraries**
```{r}
# Load required libraries
library(tidyverse) # Data manipulation and visualization
library(caret)     # For machine learning workflows
library(randomForest) # Random forest model
library(ggplot2)   # Advanced visualizations
library(DataExplorer) # Automated EDA
library(plotly)    # Interactive visualizations
```

### **Load the Dataset**
```{r}
# Load the dataset
data <- read.csv("HR-Employee-Attrition.csv")

# Inspect the dataset
head(data)
summary(data)
str(data)
```

### **Data Cleaning**
```{r}
# Check for missing values
sum(is.na(data))

# Handle missing values if any
data <- data %>% drop_na()

# Remove duplicates
data <- data[!duplicated(data), ]

# Remove zero-variance columns
zero_variance <- sapply(data, function(x) length(unique(x)) == 1)
data <- data[, !zero_variance]

# Convert categorical variables to factors
data <- data %>% mutate_if(is.character, as.factor)

# Inspect changes
str(data)
```

---

## **2. Exploratory Data Analysis (EDA)**

### **Basic EDA**
```{r}
# Automated EDA
DataExplorer::create_report(data, output_file = "EDA_Report.html")
```

### **Univariate Analysis**
```{r}
# Attrition distribution
attrition_plot <- ggplot(data, aes(x = Attrition)) + 
  geom_bar(fill = "steelblue") + 
  ggtitle("Attrition Distribution")
plotly::ggplotly(attrition_plot)

# Age distribution
age_plot <- ggplot(data, aes(x = Age)) + 
  geom_histogram(binwidth = 5, fill = "darkgreen", color = "white") + 
  ggtitle("Age Distribution")
plotly::ggplotly(age_plot)
```

### **Bivariate Analysis**
```{r}
# Attrition by Department
attrition_dept <- ggplot(data, aes(x = Department, fill = Attrition)) + 
  geom_bar(position = "dodge") + 
  ggtitle("Attrition by Department")
plotly::ggplotly(attrition_dept)

# Attrition and Overtime
overtime_attrition <- ggplot(data, aes(x = OverTime, fill = Attrition)) + 
  geom_bar(position = "dodge") + 
  ggtitle("Attrition and Overtime")
plotly::ggplotly(overtime_attrition)
```

### **Correlation Analysis**
```{r}
# Compute correlation matrix
numeric_data <- data %>% select_if(is.numeric)
corr_matrix <- cor(numeric_data)

# Melt the correlation matrix for ggplot
melted_corr <- melt(corr_matrix)

# Create the heatmap
heatmap_plot <- ggplot(data = melted_corr, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") + # Adds white gridlines between tiles
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), 
                       name = "Correlation") +
  theme_minimal() + # Minimal theme for better visualization
  theme(axis.text.x = element_text(angle = 45, hjust = 1), # Rotates x-axis labels
        axis.text.y = element_text(size = 10)) + # Adjusts y-axis text size
  labs(title = "Correlation Heatmap", x = "Features", y = "Features")

# Display the plot
heatmap_plot
```

---

## **3. Statistical Analysis**

### **Hypothesis Testing**
```{r}
# Chi-squared test for Attrition and Department
chisq.test(table(data$Attrition, data$Department))

# ANOVA for Attrition and Age
aov_result <- aov(Age ~ Attrition, data = data)
summary(aov_result)
```

---

## **4. Predictive Modeling**

### **Data Preparation**
```{r}
# Split data into training and testing sets
set.seed(123)
trainIndex <- createDataPartition(data$Attrition, p = 0.8, list = FALSE)
train <- data[trainIndex, ]
test <- data[-trainIndex, ]

# Feature scaling
preProc <- preProcess(train, method = c("center", "scale"))
train <- predict(preProc, train)
test <- predict(preProc, test)
```

### **Logistic Regression**
```{r}
# Build logistic regression model
log_model <- glm(Attrition ~ ., data = train, family = binomial)
summary(log_model)

# Predictions
log_pred <- predict(log_model, test, type = "response")
log_class <- ifelse(log_pred > 0.5, "Yes", "No")
```

### **Random Forest**
```{r}
# Build Random Forest model
rf_model <- randomForest(Attrition ~ ., data = train, ntree = 100)
print(rf_model)

# Feature Importance
importance(rf_model)
varImpPlot(rf_model)

# Predictions
rf_pred <- predict(rf_model, test)
```

### **Model Evaluation**
```{r}
# Confusion Matrix for Logistic Regression
confusionMatrix(as.factor(log_class), as.factor(test$Attrition))

# Confusion Matrix for Random Forest
confusionMatrix(as.factor(rf_pred), as.factor(test$Attrition))
```

---

## **5. Business Recommendations**

- **Key Findings**:
  - High attrition rates are observed in departments with frequent overtime.
  - Younger employees (ages 25-35) show a higher likelihood of leaving.

- **Recommendations**:
  - Introduce flexible working hours to reduce overtime pressure.
  - Create targeted retention programs for younger employees.

---

## **6. Interactive Dashboard (Optional)**
```{r}
# Define the UI for the dashboard
ui <- fluidPage(
  titlePanel("HR Attrition Analysis Dashboard"),
  
  # Layout structure
  fluidRow(
    column(4, 
           # Left panel for Key Insights
           wellPanel(
             h3("Key Insights"),
             uiOutput("insights")  # Placeholder for dynamic insights
           )
    ),
    column(4,
           # Center panel for visualization
           plotlyOutput("plot")
    ),
    column(4,
           # Right panel for dropdown menu
           selectInput("visualization", "Choose Visualization:", 
                       choices = c("Age Distribution", "Income Distribution", "Job Satisfaction", "Outliers", "Attrition by Department", "Attrition by Overtime"))
    )
  )
)

# Define server logic
server <- function(input, output, session) {
  
  # Dynamic Insights update based on selected visualization
  observeEvent(input$visualization, {
    insights_text <- switch(input$visualization,
                            "Age Distribution" = c("Younger employees (ages 25-35) tend to have higher attrition rates.",
                                                   "Attrition decreases as employees age, suggesting higher retention among older employees."),
                            "Income Distribution" = c("Employees with lower income levels show higher attrition rates.",
                                                      "Income disparity may contribute to dissatisfaction and higher turnover in lower-income brackets."),
                            "Job Satisfaction" = c("Employees with lower job satisfaction scores are more likely to leave.",
                                                   "Improving job satisfaction can significantly reduce attrition rates."),
                            "Outliers" = c("Outliers in age or income indicate exceptional cases of either early departure or long-term retention.",
                                          "Outliers may need further investigation to understand their reasons for staying or leaving."),
                            "Attrition by Department" = c("Certain departments exhibit higher attrition rates, possibly due to workload or lack of resources.",
                                                        "Departments with high stress or frequent overtime have significantly higher turnover."),
                            "Attrition by Overtime" = c("Employees working overtime are at a higher risk of attrition.",
                                                        "Reducing overtime in certain departments may help decrease attrition rates.",
                                                        "Employees not working overtime tend to have a lower attrition rate.")
    )
    
    output$insights <- renderUI({
      tags$ul(lapply(insights_text, function(text) tags$li(text)))
    })
  })
  
  # Render dynamic plot based on selected visualization
  output$plot <- renderPlotly({
    if (input$visualization == "Age Distribution") {
      # Age distribution plot
      p <- ggplot(data, aes(x = Age)) + 
        geom_histogram(binwidth = 5, fill = "lightblue", color = "white") + 
        ggtitle("Age Distribution") +
        theme_minimal()
      ggplotly(p) %>% layout(title = "Age Distribution")
      
    } else if (input$visualization == "Income Distribution") {
      # Income distribution plot
      p <- ggplot(data, aes(x = MonthlyIncome)) + 
        geom_histogram(binwidth = 1000, fill = "lightgreen", color = "white") + 
        ggtitle("Income Distribution") +
        theme_minimal()
      ggplotly(p) %>% layout(title = "Income Distribution")
      
    } else if (input$visualization == "Job Satisfaction") {
      # Job satisfaction by attrition
      p <- ggplot(data, aes(x = JobSatisfaction, fill = Attrition)) + 
        geom_bar(position = "dodge") + 
        ggtitle("Job Satisfaction by Attrition") +
        theme_minimal()
      ggplotly(p) %>% layout(title = "Job Satisfaction by Attrition")
      
    } else if (input$visualization == "Outliers") {
      # Outliers detection for age and income
      outliers_age <- function(x) {
        q1 <- quantile(x, 0.25)
        q3 <- quantile(x, 0.75)
        iqr <- q3 - q1
        lower_bound <- q1 - 1.5 * iqr
        upper_bound <- q3 + 1.5 * iqr
        x[x < lower_bound | x > upper_bound]
      }
      
      outliers_income <- outliers_age(data$MonthlyIncome)
      outliers_age_vals <- outliers_age(data$Age)
      
      p <- ggplot(data, aes(x = Age)) + 
        geom_histogram(binwidth = 5, fill = "lightblue", color = "white") + 
        geom_point(data = data[data$Age %in% outliers_age_vals, ], aes(x = Age), color = "red", size = 2) +
        ggtitle("Age Distribution with Outliers") +
        theme_minimal()
      ggplotly(p) %>% layout(title = "Age Distribution with Outliers")
      
    } else if (input$visualization == "Attrition by Department") {
      # Attrition by department
      p <- ggplot(data, aes(x = Department, fill = Attrition)) + 
        geom_bar(position = "dodge") + 
        ggtitle("Attrition by Department") +
        theme_minimal()
      ggplotly(p) %>% layout(title = "Attrition by Department")
      
    } else if (input$visualization == "Attrition by Overtime") {
      # Attrition by overtime
      p <- ggplot(data, aes(x = OverTime, fill = Attrition)) + 
        geom_bar(position = "dodge") + 
        ggtitle("Attrition and Overtime") +
        theme_minimal()
      ggplotly(p) %>% layout(title = "Attrition and Overtime")
    }
  })
}

# Run the application 
shinyApp(ui = ui, server = server)
```